---
title: "Reproducible research project 1 rework"
author: "Praveen Arora"
date: "16/11/2020"
output: html_document
---


## Assignment Instructions
1.Code for reading in the dataset and/or processing the data
2.Histogram of the total number of steps taken each day
3.Mean and median number of steps taken each day
4.Time series plot of the average number of steps taken
5.The 5-minute interval that, on average, contains the maximum number of steps
6.Code to describe and show a strategy for imputing missing data
7.Histogram of the total number of steps taken each day after missing values are imputed
8.Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends


## Step 1 reading the file and formatting the date and loading the libraries needed

```{r, echo = TRUE}

setwd("c:/users/Praveen/Desktop/Coursera/Data")
activity <- read.csv("activity.csv")
library(data.table)
library(lubridate)
library(lattice)
```

##displaying the loaded data

```{r,echo=TRUE}

dim(activity)
head(activity)
names(activity)
summary(activity)
#changing to date format

activity$date<- ymd(activity$date)

#total number of missing data
sum(is.na(activity$steps))


```

## Step 2 plotting the histogram of number of steps in a day and seeing if it comes in .MD file

```{r,echo=TRUE}

daysum <- aggregate(steps ~ date, activity, sum, na.rm=TRUE)
hist(daysum$steps) 
```


## Step3 calculating mean, median, sum by date and by interval and plotting histogram of daysteps

```{r,echo=TRUE}
meanstepsinaday <- mean(daysum$steps)
meanstepsinaday

medianstepsinaday <- median(daysum$steps)
medianstepsinaday
```

##creating variables to cover number of steps within interval, additional variable created for testing functionality later

```{r,echo=TRUE}

intervalsum <- aggregate(steps ~ interval, activity, sum, na.rm=TRUE)
intervalmean <- aggregate(steps ~ interval, activity, mean, na.rm=TRUE)
intervalmedian <- aggregate(steps ~ interval, activity, median, na.rm=TRUE)

```
## Step 4, plotting average steps in an interval
```{r,echo=TRUE}
plot(steps~interval,data=intervalmean,type="l")
```

## Step 5, interval with max steps
```{r,echo=TRUE}
intervalmaxsteps <- intervalmean[which.max(intervalmean$steps),]$interval
intervalmaxsteps
```

## Step 6, total missing values to replaced by mean value for that interval

```{r,echo=TRUE}
missingvaluetotal <- sum(is.na(activity$steps))
missingvaluetotal
```

###this function gets the meanvalue for the specific interval
```{r,echo=TRUE} 
meanstepsperinterval <- function(interval)
{
    intervalmean[intervalmean$interval==interval,]$steps
}

```

###this function looks at every row, wherever missing data is found, it obtains mean value for that interval & substitutes it)
###The new data set with no missing values is contained in the variable activitynomissing

```{r,echo=TRUE}

activitynomissing <- activity
for (i in 1:nrow(activitynomissing))
{
	if(is.na(activitynomissing[i,]$steps))
	{
		activitynomissing[i,]$steps <- meanstepsperinterval(activitynomissing[i,]$interval)
	}
}
```


## Step 7 Making a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 

```{r,echo=TRUE}

totalstepsperdaynomissing <- aggregate(steps~date,data=activitynomissing,sum)

hist(totalstepsperdaynomissing$steps)

meannomissing <- mean(totalstepsperdaynomissing$steps)

mediannomissing <- median(totalstepsperdaynomissing$steps)

meannomissing

mediannomissing

```

###the median differs from the estimate 
###since mean value used to substiture NA data, mean is the same as before

## As an addition step, I used another direct replacement without creating intermediate function, tested with sum, max, mean and median
#####intervalmax <- aggregate(steps ~ interval, activity, max, na.rm=TRUE)

#####activitynomissing1 <- activity	
#####for (i in 1:nrow(activitynomissing))
#####{
#####if(is.na(activitynomissing1[i,]$steps))
#####{
#####activitynomissing1[i,]$steps <- intervalmax[intervalmax$interval==(activitynomissing[i,]$interval),]$steps
#####}
#####}

#####totalstepsperdaynomissing1 <- aggregate(steps~date,data=activitynomissing1,sum)

#####meannomissing1 <- mean(totalstepsperdaynomissing1$steps)

#####mediannomissing1 <- median(totalstepsperdaynomissing1$steps)

#####meannomissing1

#####mediannomissing1


## Step 8. Creating factor variable on the weekday... using for loop to create factor variable for each row 

```{r,echo=TRUE}

activitynomissing$date <- ymd(activitynomissing$date)

activitynomissing$day <- weekdays(activitynomissing$date)

for (i in 1:nrow(activitynomissing))
{
	if(activitynomissing[i,]$day %in% c("Saturday","Sunday"))
	{
		activitynomissing[i,]$day="weekend"
	}
	else
	{
		activitynomissing[i,]$day="weekday"
	}
}

```

##making a panel plot containing a time series plot of the 5-minute interval (x-axis) and 
##the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 
 
```{r,echo=TRUE}

stepbydaytype <- aggregate(activitynomissing$steps~activitynomissing$interval + activitynomissing$day,activitynomissing,mean)

names(stepbydaytype) <- c("interval", "day", "steps")

xyplot(steps~interval|day,stepbydaytype,type="l",layout=c(1,2),xlab="Interval",ylab="# Steps")
```
